---
title: "Fisheries ANOVA exercise"
author: "N.M. Baran, D. Thurtle-Schmidt"
date: "2024-10-01"
output: pdf_document
---


Set your working directory and read in the fisheries data:

```{r}
fisheries <- read.csv("Hartvigsen_FishData.csv", header = TRUE)
```

Q1: Write your null and alternative hypotheses for conducting the ANOVA:
    
    H0: mu(low) = mu(med) = mu(high)
    HA: At least one mean is different


Next, we need to check our assumptions.

Check for normality:
```{r}
boxplot(fisheries$Low, fisheries$Med, fisheries$High)
```
```{r}
shapiro.test(fisheries$Low)
shapiro.test(fisheries$Med)
shapiro.test(fisheries$High)
```
Check the equality of variances:
```{r}
var(fisheries$Low)
var(fisheries$Med)
var(fisheries$High)
```
Q2: Do our assumptions for ANOVA appear to be satisfied? Explain.

A2: Yes. Each of the variables are normally distributed and the variance is not greater than 4-fold between the 3 variables.

In order to conduct our analysis, we need to prepare the data for anova. This code converts the data frame from wide to long format:
```{r}
fish_stack <- stack(fisheries)
names(fish_stack) <- c("Mass", "Treatment")
```

Q3: How did this code change/reshape the data frame? (Note that getting your data into the right format is one of the harder/more annoying parts of stats!)

A3: The data frame is now organized with the growth of the fish (Mass) for the three different food supplementation rates in a single column. The column of Treatment lists the Treatment for each condition.

Now run the ANOVA:
```{r}
fish.aov <- aov(fish_stack$Mass~fish_stack$Treatment) #Runs the anova and stores it as an object
fish.aov #Provides only sum of squares, residuals, and dfs...
summary(fish.aov) #Provides more details, including F-value and p-value!
```

Showing that the F value is the *ratio of the mean squares* and that using the F-distribution you can find the p-value with the degrees of freedom indicated.
```{r}
245.03/16.84
1-pf(14.55, 2, 15) #Here we're using the F-distribution!
```
Q4: Interpret the meaning of this ANOVA. Do we reject or fail to reject our null hypothesis?

We can reject our null hypothesis because the p-value = 0.000308. This suggests that at least one of the three groups is different from one of the other ones.

We're not done yet! We don't know which group(s) are different from the others. Thus, we need to perform a posthoc analysis:
```{r}
TukeyHSD(fish.aov) #note that you input the anova object that you created above!
```
Q5: Which pairwise comparisons are statistically significant?

A: The Low conditions is significantly different from both the Medium and High conditions. 
