---
title: "Probability exercises"
author: "SMWadgymar, NMBaran"
date: '2022-08-16'
output:
  pdf_document: default
  html_document: default
---

####Exercises

1) Simulating coin flips:  Generate 10 separate random flips with probability 0.7 of producing heads.

```{r}
rbinom(10, 1, .7)
```

Q: What kind of values do you see and what do they represent?

A: Each number is 1 or 0 depending on whether the coin was heads or tails. 


2) Simulating draws from a binomial distribution:  Generate 100 occurrences of flipping 10 coins, each with 70% probability of producing heads. Produce a table of the probability distribution that you generate (hint: your table should report probabilities and not counts).

```{r}
flips<-rbinom(100, 10, 0.7)

prop.table(table(flips))

#Could also use table(flips)/100 to get the table
```

Q: What kind of values do you produce and what do they represent?  Plot a histogram and describe the shape that your probability distribution takes.  What does the freq=F argument mean?

A: When you use the `freq=F' argument in hist(), you are asking for the relative frequency histogram, which has total area 1. Area is proportional to relative frequency. 

```{r}
hist(flips, freq=F)
```



3) Calculate the exact probability that 2 heads will arise from 10 coin flips with a 70% probability of coming up tails.  

```{r}
dbinom(2, 10, 0.3) 
#Note that the word problem lists the probability of getting tails
```

Q: Compare your answer with a simulation of 10,000 trials.  Do the two approaches yield similar results?

A: 
```{r}
mean(rbinom(10000, 10, 0.3)==2)
```
These are similar, but not identical!


4) Calculate the cumulative probability that at least five coins out of 10 are heads with a 30% probability of coming up heads.  

```{r}
1-pbinom(4, 10, 0.3)
pbinom(4, 10, 0.3, lower.tail=FALSE)
##Note that we have to subtract from 1 to get the right range
```

Q: Compare your answer with a simulation of 10,000 trials.  Do the two approaches yield similar results?

A: These are similar, but not identical!

```{r}
mean(rbinom(10000, 10, 0.3)>= 5)
```



5) Repeat the simulation you ran in exercise (3) where you estimated the exact probability that two heads will occur with 10, 100, 1,000, 10,000, and 100,000 trials.  Produce a plot depicting the number of trails on the x-axis and the associated probabilities you calculated on the y-axis.  

Hint: The c() function lets you combine values into a vector or list.  You will need to make a vector of probabilities for the y axis using and a vector of your sample sizes for the x axis using c(#, #, #, #, #).

```{r}
r10<-mean(rbinom(10, 10, 0.3)==2)
r100<-mean(rbinom(100, 10, 0.3)==2)
r1000<-mean(rbinom(1000, 10, 0.3)==2)
r10000<-mean(rbinom(10000, 10, 0.3)==2)
r100000<-mean(rbinom(100000, 10, 0.3)==2)

r<-c(r10, r100, r1000, r10000, r100000)
n<-c(10, 100, 1000, 10000, 100000)

plot(n, r, xlab="Number of replicates", ylab="Probability of 70% flips coming up tails")
```

Produce a second plot but this time log-transform the number of trials.  Make sure to adjust your axis labels appropriately.  

```{r}
plot(log(n), r, xlab="Log of number of replicates", ylab="Probability of 70% flips coming up tails")
```

Q: Which simulation yields a result most similar to the exact probability?  What pattern do you see in your first graph?  What does your second graph reveal that your first graph did not?

A: The simulation with the largest sample size yields a result most similar to the exact probability.  The sample sizes differ by orders of magnitude, which makes a log scale more appropriate for graphing.  Once you log-transform the number of replicates, you can see that sample sizes of 100 and larger start to converge on the same value (the exact probability, which you calculated in question 3 using dbinom(2, 10, 0.3)).



####Background: joint events

6) Let's simulate two events, A and B.

```{r}
A<-rbinom(100000, 1, 0.5)
B<-rbinom(100000, 1, 0.5)
```

Q: Describe these events.

A: These events represent a single draw of 100000 fair coins that are flipped once.



7) Q: What is the probability of event A occurring (that is, equaling 1)?  How would you calculate that in R?  What about the probability of event B occurring?

A:
```{r}
sum(A)/100000

sum(B)/100000
```




8) We can implement the General Addition Rule in R this way, where the | symbol means 'or':

```{r}
mean(A|B)
```


We can implement the Multiplication Rule for Independent Events in R this way, where the & symbol means 'and':

```{r}
mean(A&B)
```


If events A and B are independent, and A has a 50% chance of happening and B has a 30% chance of happening, what is the probability that they will both happen?  Use simulations to answer this question.

```{r}
A<-rbinom(1000, 1, 0.5)
B<-rbinom(1000, 1, 0.3)
mean(A & B)
```


9) Expanding on exercise (8), event C has a 70% chance of happening.  What is the probability that events A, B, and C all happen?

```{r}
C<-rbinom(1000, 1, 0.7)
mean(A & B & C)
```


10) If events A and B are independent, and A has a 40% chance of coming up heads and B has a 75% chance of coming up heads, what is the probability that either A or B will come up heads?  Use simulations to answer this question.

```{r}
A<-rbinom(1000, 1, 0.4)
B<-rbinom(1000, 1, 0.75)
mean(A|B) 
```


11) Suppose X is a random binomially distributed variable (10, 0.3) and Y is another random binomially distributed variable (10, 0.65), and that they are independent.  What is the probability that either of these variables is less than or equal to 5?  Estimate this probability both using simulations of 10,000 trials and by calculating exact cumulative densities.  

Using simulations:
```{r}
X<-rbinom(100000, 10, 0.3)
Y<-rbinom(100000, 10, 0.65)
mean(X<=5|Y<=5)
```

Calculating exact cumulative densities:
```{r}
prob_X_less<-pbinom(5, 10, 0.3)
prob_Y_less<-pbinom(5, 10, 0.65)
prob_X_less + prob_Y_less - prob_X_less*prob_Y_less
```



Q: How do these two approaches compare?

A: They are extremely similar because we had large sample sizes in our simulation.
