---
title: "Multiple Linear Regression"
author: "S. Wadgymar, N.M. Baran"
date: "10/27/2024"
output:
  pdf_document:
    latex_engine: xelatex
---
*adapted from OpenIntro biostats chapter 7 labs

In most practical settings, more than one explanatory variable is likely to be associated with a response. Multiple linear regression is an extension of simple linear regression that allows for more than one predictor variable in a linear model. As with simple linear regression, the response variable must be numerical, but the predictor variables can be either numerical or categorical. 

There are several applications of multiple regression. One of the most common applications in a clinical setting is estimating an association between a response variable and primary predictor of interest while adjusting for possible confounding variables.

First, we will explore multiple regression model by examining the possible association between cognitive function and the use of statins after adjusting for a potential confounding variable, age.

Reminder: Statins are a class of drug widely used to lower cholesterol. Research suggests that adults with elevated low density lipoprotein (LDL) cholesterol may be at risk for adverse cardiovascular events. A set of guidelines released in 2013 recommended statin therapy in individuals who are at high risk of adverse cardiovascular events, including individuals with Type II diabetes and moderately high LDL and non-diabetic individuals with atherosclerotic cardiovascular disease and high LDL. If these guidelines were to be followed, almost half of Americans ages 40 to 75 and nearly all men over 60 would be prescribed a statin.

However, some physicians have raised the question of whether treatment with a statin might be associated with an increased risk of cognitive decline.

This lab uses data from the Prevention of REnal and Vascular END-stage Disease (PREVEND) study. Clinical and demographic data for 4,095 individuals are stored in the PREVEND dataset in the oibiostat package. 

```{r}
library(oibiostat)
data(prevend.samp)
```

If the above code doesn't work, you need to install the oibiostat package using the following commands (minus the #)
```{r}
#install.packages("devtools")
#devtools::install_github("OI-Biostat/oi_biostat_data")
```


################################
1) Statin use is coded as an integer vector, where 0 represents a non-user and 1 represents a user. Run the code chunk below to convert the variable Statin into a factor variable, with levels NonUser and User. Describe what this code does below the code chunk (look at the data before and after running it).


```{r}
prevend.samp$Statin <- factor(prevend.samp$Statin, levels = c(0,1), labels = c("Nonuser", "User") )
```

A: This line of code recodes the integer dummy variable for statin use (0/1) and labels 0 to 'Nonuser' and 1 to 'User'.

2. Explore the relationship between RFFT score and statin use with the data in prevend.samp

a) Statin use is coded as an integer vector, where 0 represents a non-user and 1 represents a user. Run the code chunk below to convert the variable Statin into a factor variable, with levels NonUser and User. Talk through what this code does (look at the data before and after running it).

```{r}
prevend.samp$Statin <- factor(prevend.samp$Statin, levels = c(0,1), labels = c("Nonuser", "User") )
```


b) Create a plot showing the association between RFFT score and statin use. Describe what you see.

Since statin use is a categorical variable, we know that the relationship is linear (the relationship between any two points is linear). Thus, it is best to look at the data with a boxplot:

```{r}
boxplot(prevend.samp$RFFT~prevend.samp$Statin)
```

It looks like there is possibly a relationship between statin use and lower (negative) RFFT score.


################################
2) Adjust the code in the code chunk below to generate a scatterplot of RFFT score by Age.  Note that R will assign colors alphabetically to levels in alphabetical order.  Black comes before red alphabetically and Nonuser comes before User, therefore black represents Nonusers and red represents users.

```{r}
plot(prevend.samp$Age, prevend.samp$RFFT,
     pch = 19, col = prevend.samp$Statin, 
     xlab = "Age (yrs)",
     ylab = "RFFT Score",
     main = "")

legend('topright', legend=c("Nonuser", "User"), fill=c("black", "red"), bty='n')
```


Let's look at the correlation between Age and RFFT score.  Interpret the correlation coefficient below.

```{r}
cor(prevend.samp$Age, prevend.samp$RFFT)
```

A: This correlation describes a negative, moderate association between age and RFFT score. 

################################

3) Let's fit a simple regression model and interpret the slope coefficient. The lm function stands for linear model.  The arguments are Y~X.  Examine the output of the model and interpret all of its elements.  

```{r}
statin_lm <- lm(prevend.samp$RFFT~prevend.samp$Statin)
summary(statin_lm)
```

A: The first thing we should look at here is the F-statistic and p-value, which tells us that at least one of our explanatory variables (here we only have Statin use) is associated with the response variable (RFFT scores). 

The estimate for the intercept is 70.714 and it is statistically significant (p<0.05). The estimate for the effect of statin ue is -10.053 and this is significant, as well (P<0.05). If we build our mode in the form of y=b0 + b1(StatinUse), where b0 is the intercept and b1 is the slope of the line, the equation is:

y = 70.714-10.053(StatinUse). 

Thus the predicted RFFT score for a nonUser of statin is 70.714 -10.053 (0) = 70.714. The predicted RFFT score for a user of statin is 70.714-10.054(1) = 60.66. 

The adjusted R-squared is 0.02194, which means that 2.194% of the variation in RFFT scores can be explained by statin use. 

Q: Discuss whether the model from part 3) is sufficient for understanding whether statin use is associated with decreased cognitive ability.

A: Although the p-value for the slope is very small, we still likely want to be cautious about this relationship as the decline in age could simply be due to those people that take statin tend to be older. 

Age is a potential confounder for the relationship between statin use and cognitive function. If older participants tend to use statins, and higher age is associated with lower cognitive ability, perhaps the observed negative association between cognitive ability and statin use is primarily driven by age.


################################

4) Age is a potential confounder for the relationship between statin use and cognitive function. If older participants tend to use statins, and higher age is associated with lower cognitive ability, perhaps the observed negative association between cognitive ability and statin use is primarily driven by age.

Let's fit a multiple regression model for predicting RFFT score from statin use and age. Examine the output of the model and interpret all of its elements.  

```{r}
model.RFFTvsStatinAge = lm(prevend.samp$RFFT~prevend.samp$Statin + prevend.samp$Age)
summary(model.RFFTvsStatinAge)
```
a) Write the equation of the linear model.
RFFT = .8509*statin_use + -1.271*age + 137.882

b) Interpretation of intercept:

The line of best fit intersects the Y-axis (i.e. when X=0) at 137.8822. This estimate is significant at P<0.05.

c) Interpretation of slope for Statin: 

The estimate for statin is 0.8509, but it is not significant at alpha<0.05. If age is kept constant than the RFFT score increases by .8509 for someone who uses statin compared to someone that doesn't.

d) Interpretation of slope for Age: 

The estimate for the effect of age on RFFT score is -1.2710 and the effect is highly significant. If statin use is kept constant, then the RFFT score is expected to decrease by 1.271 as age increases by a year.

e) Interpretation of R2: 

The model explains 28.23% of the variance in RFFT scores. 



################################

5) b) As with simple linear regression we need to check assumptions before interpreting our model. Check all assumptions of the linear model. (hint: to check for linearity assess the residuals against each of the predictors. To check for constant variance check the residuals against the response variable.) Run the code below and address whether assumptions are met.

```{r}
#Extract residuals
hist(resid(model.RFFTvsStatinAge))

#Examine residuals (here, fitted values = predicted values)
#Just look at the first two plots for now
plot(model.RFFTvsStatinAge)

#Variance test of residuals against Statin use
var.test(resid(model.RFFTvsStatinAge)~prevend.samp$Statin)
```

A: The residuals look pretty normal, so this seems okay (some departures from normality at the tails). There are no apparent trends; the data scatter evenly above and below the horizontal line. There does not seem to be remaining nonlinearity with respect to age after the model is fit.

There is some issue potentially with the continuity of variance, but it is not an extreme issues. Variance within the statin groups looks okay. 

The linearity between age and RFFT score looks good. Note that it is not necessary to assess linearity with respect to statin use since statin use is measured as a categorical variable. A line drawn through two points (that is, the mean of the two groups defined by a binary variable) is necessarily linear.

Are the observations independent? It is reasonable to assume that the observations in this dataset are independent. The participants were recruited from a large city in the Netherlands for a study focusing on factors associated with renal and cardiovascular disease.

################################

6) How does the simple linear model compare to the multiple regression?  Answer the questions below:

a) The adjusted R2 value accounts for the increased number of predictors and should be compared across models to determine the best model. How does the adjusted R2 of the simple linear model compare with the multiple linear model? Which do you think is a better model for the data?

A: The adjusted R2 value is much higher for the multiple regression (0.2823 vs 0.02194), which means a greater proportion of the variance is explained. The second model is better because a greater proportion of the variance is explained.


b) As in simple linear regression, inferences can be made about the slope parameters estimated by the model slope coefficients. Based on the multiple regression model, is there a statistically significant association between RFFT score and statin use? 

A: No. In the simple linear model, statin use was a significant predictor of RFFT score, but this effect disappeared when we included age in the model (i.e. when we 'controlled' for age). This suggests that statin use and age were confounding variables. Once age is accounted for there is not enough evidence to reject the null hypothesis that there is no association (beta = 0) between statin use and RFFT.



c) In a clinical setting, the interpretive focus lies on reporting the nature of the association between the primary predictor and the response, while specifying which potential confounders have been adjusted for. Briefly respond to a clinician who is concerned about a possible association between statin use and decreased cognitive function, based on the analyses conducted in this lab.

A: Once you adjust for the confounding variable of age, there is no evidence to suggest that statin use is associated with a lower RFFT score. Specifically after adjusting for age, the estimated difference in mean RFFT score between statin users and nonusers is .85 points; there is a 74% chance of observing such a difference if there is no difference between mean RFFT score in the population of statin users and nonusers.

        
###########################

Now let's look at if there is an interaction between statin use and age on RFFT score. Look at the two formulations of the model below:

```{r}
model.RFFTxStatinAge = lm(prevend.samp$RFFT~prevend.samp$Statin + prevend.samp$Age+prevend.samp$Statin*prevend.samp$Age)
summary(model.RFFTxStatinAge)
```        
```{r}
model.RFFTxStatinAge = lm(prevend.samp$RFFT~prevend.samp$Statin*prevend.samp$Age)
summary(model.RFFTxStatinAge)
```  

Q: Is there evidence of significant interaction between age and statin use on RFFT scores?

A: No. Since the interaction term is not significant, the effect of age on RFFT score does not depend on age. Note also that the adjusted R2 is the same (i.e. has not increased at all)!


################################
7) Let's add the lines of best fit predicted by the models to the graph. We'll use the multiple regression without the interaction term. 

First, let's identify the slopes and intercepts.  The coef function will isolate the coefficients (parameters) from the model.  There are three parameters from this model.  You can list individual parameters by specifying [1], [2], or [3] at the end of the code (I do that in the next code chunk).

```{r}
coef(model.RFFTvsStatinAge)
```


Next, let's add some trend lines.  The abline(a,b) function will let you specify the intercept (a) and the slope (b) for a line.  I wrote the function to graph the line for Statin non-users.  You need to adjust the code to graph the line for Statin users.  Note that you have to run the entire code chunk to see if the lines are working (you can't just run the abline code by itself).

```{r}
##Paste the code for the graph you previously made

plot(prevend.samp$Age, prevend.samp$RFFT,
     pch = 19, col = prevend.samp$Statin, 
     xlab = "Age (yrs)",
     ylab = "RFFT Score",
     main = "")


##line for Statin non-users
abline(a=coef(model.RFFTvsStatinAge)[1], 
       b=coef(model.RFFTvsStatinAge)[3],
       lwd=2)

##line for Statin users
abline(a=coef(model.RFFTvsStatinAge)[1], 
       b=coef(model.RFFTvsStatinAge)[3],
       lwd=2, lty=2, col="red")

```
